|

### 2024-06-30 manuscript.Rmd
Incorporated co-author comments and questionnaire data. Anonymised for submission.

---
title: |
  | Neural correlates of the uncanny valley effect
  | for robots and hyper-realistic masks 
author: |
  | Shona Fitzpatrick$^1$, Ailish K. Byrne$^2$, Alex Headley$^1$, Jet G. Sanders$^3$,
  | Helen Petrie$^4$, Rob Jenkins$^1$ & Daniel H. Baker$^{1,5}$
output:
  bookdown::pdf_document2:
    fig_caption: yes
    toc: no
    keep_tex: yes
  word_document: default
  pdf_document:
    toc: no
  html_document: default
bibliography: references.bib
csl: elife.csl
---

1. Department of Psychology, University of York, York, UK
2. School of Medicine, Keele University, Newcastle-under-Lyme, Staffordshire, UK
3. Department of Psychological and Behavioural Science, London School of Economics, London, UK
4. Department of Computer Science, University of York, York, UK
5. Corresponding author, email: daniel.baker\@york.ac.uk

@@

processdata <- 2  # this flag determines the amount of processing, with 3 levels:
# 0 - do no processing, generate the pdf using existing versions of all figures
# 1 - generate figures using the processed group data (requires 120MB of storage)
# 2 - do statistics and bootstrapping using processed data

@@

# check which packages are installed, install the missing ones, and activate
packagelist <- c('knitr','e1071','signal','osfr','tictoc','BayesFactor','reshape2','powerplus','R.matlab') # list of CRAN packages
missingpackages <- packagelist[!packagelist %in% installed.packages()[,1]]
if (length(missingpackages)>0){install.packages(missingpackages)}
toinstall <- packagelist[which(!packagelist %in% (.packages()))]
invisible(lapply(toinstall,library,character.only=TRUE))

@@

  if (!file.exists('local/channel.mat')){
  osfproject <- osf_retrieve_node('uzbns')
  osffiles <- osf_ls_files(osfproject)
  fid <- which(osffiles$name=='channel.mat')
  osf_download(osffiles[fid,], path='local/',progress=TRUE)}

  if (!file.exists('local/channeldata.csv')){
  osfproject <- osf_retrieve_node('uzbns')
  osffiles <- osf_ls_files(osfproject)
  fid <- which(osffiles$name=='channeldata.csv')
  osf_download(osffiles[fid,], path='local/',progress=TRUE)}

@@

Viewing artificial objects and images that are designed to appear human can elicit a sense of unease, referred to as the 'uncanny valley' effect. Here we investigate neural correlates of the uncanny valley, using still images of androids (robots designed to look human), and hyper-realistic silicone masks, in two independent experiments. In both cases, human-like stimuli were harder to distinguish from real human faces than examples that were clearly not designed to mimic humans (mechanical robots, and Halloween masks). Stimulus evoked potentials did not show convincing differences between faces and either androids or realistic masks when using traditional univariate statistical tests. However, a more sensitive multivariate analysis identified two regions of above-chance decoding, indicating neural differences in the response between human faces and androids/realistic masks. The first time window was around 100-200ms post stimulus onset, and most likely corresponds to low-level image differences between conditions. The second time window was around 600ms post stimulus onset, and may reflect top-down processing, and perhaps correspond to the subjective sense of unease characteristic of the uncanny valley effect. Objective neural components might be used in future to rapidly train generative artificial intelligence systems to produce more realistic images that are perceived as natural by human observers.

@@

Many people report an aversion to entities that are superficially human-like, but on closer inspection are actually artificial. Examples include humanoid robots (androids), puppets, hyper-realistic masks, and computer-generated images or movies. The term 'uncanny valley' [@Mori1970; English translation in @Mori2012] describes the idea that entities that are clearly human or clearly artificial do not evoke unease, whereas artificial entities that are human-like are disconcerting. Understanding these experiences is increasingly important as artificial entities become more integrated into our everyday lives, however at present relatively little is known about the neural underpinnings of the uncanny valley effect [for a recent review, see @Vaitonyte2023].

Neural responses to faces and bodies in general are well-characterised, and there appear to be specialised brain regions devoted to both [reviewed in @Hu2020]. For example, areas of the occipital lobe [@Gauthier2000] and fusiform gyrus [@Kanwisher1997] respond more to faces than non-face stimuli, and sections of extrastriate cortex are responsive to bodies [@Downing2001]. There are also event-related potential (ERP) signals associated with face and body stimuli, though their precise purpose is still debated [@Thierry2007]. It seems highly likely that 'uncanny' images will activate these same processes, yet it is unclear whether the sense of unease they produce occurs at bottom-up sensory stages, or is modulated by more top-down cognitive factors.

One previous study by @Saygin2012 measured fMRI responses to moving stimuli designed to elicit an uncanny valley effect. They found repetition suppression effects in action-specific brain regions responding to movies of androids that had a biological appearance, but mechanical motion. These effects were stronger than for movies of humans or mechanical robots performing the same actions. A more recent electroencephalography (EEG) study [@Urgen2018] identified a difference in the N400 component between dynamic and static conditions using the same stimuli. Although this difference was strongest over frontal electrodes, source reconstruction of the N400 itself suggested a left-lateralised source in temporo-parietal cortex, consistent with the fMRI results [@Saygin2012]. The authors interpret both of these findings as being due to the discrepancy between the human-like appearance and the clearly non-biological motion of the robot.

Our aim here was to further investigate neural correlates of the uncanny valley effect. We achieve this through two EEG experiments, in which we measure neural responses to static images. In the first experiment, the stimuli were humans, machine-like robots, and human-like robots (see Figure \ref{fig:stimexamples}a). In the second experiment we aimed to generalise the finding by using images of people wearing no masks, wearing obvious masks (e.g. carnival or Halloween masks), and wearing hyper-realistic silicone masks [@Sanders2017] (see Figure \ref{fig:stimexamples}b). Rather than focus on specific ERP components, we use a multivariate pattern classification approach to identify time windows in which information in the EEG signal can be used to distinguish between pairs of conditions. Our rationale is that timepoints where signals evoked by human faces can be distinguished from those evoked by human-like robots, or hyper-realistic masks, are candidates for a neural signature of the uncanny valley effect.

```{r stimexamples, fig.cap="Illustrative stimuli from the same categories as used in Experiments 1 and 2. Row (a) shows a human face, an android and a robot, all against white backgrounds. Row (b) shows a human face, a hyper-realistic silicone mask, and a carnival mask, all against natural backgrounds. Images were taken from a variety of sources that permit reuse in academic contexts and in most cases were not part of the stimulus set from the experiments. The silicone mask image was taken by the authors, and was used in Experiment 2.", fig.align="center", out.width='80%', echo=FALSE}

@@

A total of 29 participants completed Experiment 1 (12 male, 17 female), and 30 participants completed Experiment 2 (7 male, 23 female). None of the participants had previously taken part in a study using these stimuli, and all were naïve to the hypotheses and wore their normal optical correction if required. Written informed consent was collected before each experiment began, and all procedures were approved by the Ethics committee of the Department of Psychology at the University of York.

@@

In Experiment 1, the stimulus set consisted of a total of 90 images, evenly split between three categories: real faces, human-like robots, and mechanical robots. Images all showed the head and shoulders of the subject, had white backgrounds, and were sourced from the internet. In Experiment 2, the stimulus set consisted of a total of 296 images, comprising real faces (148 images), people wearing silicone masks (74 images), and people wearing obvious masks of the sort typically worn for carnivals and Halloween celebrations (74 images). The backgrounds of these images were more heterogeneous, and showed the natural surroundings of the subject. In both experiments, images involved examples of both genders, and of varied ethnic backgrounds.

@@

### Experiment 1: robots

@@

### Experiment 2: hyper-realistic masks

Participants were shown all 296 images in a random order in each of three blocks. In the first block, stimuli subtended $5.5\times7.5$ degrees of visual angle when viewed at a distance of 57cm. In the second block, stimuli doubled in size (width and height), and subtended $11\times15$ degrees at the same viewing distance. In the third block, stimuli doubled in size again, and subtended $22\times30$ degrees. The rationale for the size manipulation was to investigate whether increasing levels of detail made the silicone masks more identifiable [@Sanders2017], however as that is not the main focus of the current paper we collapse results across size conditions. Stimuli were presented for 250ms, and participants indicated whether they thought each image contained a real face or a mask, using a two-button trackball. The button assignment (whether the left button indicated a face or a mask, and vice versa) was determined randomly for each participant, but remained constant throughout the whole experiment. Text reminding the participant of the button assignment was present continuously in the lower right corner of the screen, far from the area of the screen where the stimuli were presented. A central fixation cross was also present throughout. After each response there was a random duration blank period with a mean duration of 1000ms and a standard deviation of 200ms. Each block lasted around 8 minutes.

@@

EEG signals were recorded during each block, and saved to disc for subsequent offline analysis. We used components of the EEGlab toolbox [@Delorme2004] to import the data into Matlab and collate data across blocks. We then used Brainstorm [@Tadel2011] to filter the data using a bandpass filter (0.5 to 30Hz), epoch by condition, and subtract a pre-trial baseline (the mean voltage for the 200ms before stimulus onset). Five participants were excluded from the EEG analysis of each experiment due to excessive noise. Our attempts to clean up the data from these participants using independent components analysis were unsuccessful. Their behavioural data were still included in the analysis. In Experiment 2 we combined data across all three image sizes, as this was not the main focus of our analysis.

@@

Multivariate pattern analysis was conducted by training a linear support vector machine algorithm (LibSVM; @Chang2011) to discriminate between patterns of activity across electrodes at a specific time point. The patterns came from the human face condition and one of the other conditions, for a single participant. Four examples of each pattern were calculated by averaging over random subsets of 20\% of the available trials from a given condition, and using these to train the classifier. The accuracy of the classifier was tested on the averaged remaining trials (that were not used in training) for each condition. This process was repeated 1000 times with different trial permutations to obtain an average accuracy, where chance performance is at 50\% correct. The analysis was carried out at all time points, and also for all participants. We then averaged classifier accuracy across participants, and calculated one sample Bayesian t-tests comparing to chance performance at each time point as described above.

@@

Raw data, processed data, and analysis scripts are freely available through the project repository at: https://doi.org/10.17605/OSF.IO/5NZ2H

@@

@@

EEG activity showed a clear visually evoked potential over posterior electrode sites (see Figure \ref{fig:RobotsData}c), with typical components found in response to visual stimuli (the P100, N170 and P200 are indicated in the figure). Pairwise comparisons of conditions are shown in Figure \ref{fig:RobotsMVPA}a,b. In general there is a tendency for the ERP response to human faces to diverge slightly from the other two conditions, however the evidence for a convincing difference was not compelling. Bayes factors exceeded 3 for only a small number of time points around 300-400ms in the comparison between human and robot images (see yellow bars at y = -8 in Figure \ref{fig:RobotsMVPA}a), but these differences were small considering the variance in the data.

@@

@@

>>>>

---
title: |
  | Neural correlates of the uncanny valley effect
  | for robots and hyper-realistic masks 
output:
  bookdown::pdf_document2:
    fig_caption: yes
    toc: no
    keep_tex: yes
  word_document: default
  pdf_document:
    toc: no
  html_document: default
bibliography: references.bib
csl: computers-in-human-behavior.csl
---

[**Author information redacted for blind peer review**]

<!-- author: | -->
<!--   | Shona Fitzpatrick$^1$, Ailish K. Byrne$^2$, Alex Headley$^1$, Jet G. Sanders$^3$, -->
<!--   | Helen Petrie$^4$, Rob Jenkins$^1$ & Daniel H. Baker$^{1,5}$ -->

<!-- 1. Department of Psychology, University of York, York, UK -->
<!-- 2. School of Medicine, Keele University, Newcastle-under-Lyme, Staffordshire, UK -->
<!-- 3. Department of Psychological and Behavioural Science, London School of Economics, London, UK -->
<!-- 4. Department of Computer Science, University of York, York, UK -->
<!-- 5. Corresponding author, email: daniel.baker\@york.ac.uk -->

@@

processdata <- 0  # this flag determines the amount of processing, with 3 levels:
# 0 - do no processing, generate the pdf using existing versions of all figures
# 1 - generate figures using the processed group data (requires 120MB of storage)
# 2 - do statistics and bootstrapping using processed data

@@

# check which packages are installed, install the missing ones, and activate
packagelist <- c('knitr','e1071','signal','osfr','tictoc','BayesFactor','reshape2','R.matlab') # list of CRAN packages
missingpackages <- packagelist[!packagelist %in% installed.packages()[,1]]
if (length(missingpackages)>0){install.packages(missingpackages)}
toinstall <- packagelist[which(!packagelist %in% (.packages()))]
invisible(lapply(toinstall,library,character.only=TRUE))

@@

filelist <- c('channel.mat','channeldata.csv','maskquestionnairelong.csv','robotquestionnairelong.csv')

for (f in 1:length(filelist)){
  if (!file.exists(paste0('local/',filelist[f]))){
  osfproject <- osf_retrieve_node('uzbns')
  osffiles <- osf_ls_files(osfproject)
  fid <- which(osffiles$name==filelist[f])
  osf_download(osffiles[fid,], path='local/',progress=TRUE)}
}

@@

Viewing artificial objects and images that are designed to appear human can elicit a sense of unease, referred to as the 'uncanny valley' effect. Here we investigate neural correlates of the uncanny valley, using still images of androids (robots designed to look human), and humans wearing hyper-realistic silicone masks, as well as still images of real humans, in two experiments. In both experiments, human-like stimuli were harder to distinguish from real human faces than stimuli that were clearly not designed to mimic humans but contain facial features (mechanical robots and Halloween masks). Stimulus evoked potentials did not show convincing differences between faces and either androids or realistic masks when using traditional univariate statistical tests. However, a more sensitive multivariate analysis identified two regions of above-chance decoding, indicating neural differences in the response between human faces and androids/realistic masks. The first time window was around 100-200ms post stimulus onset, and most likely corresponds to low-level image differences between conditions. The second time window was around 600ms post stimulus onset, and may reflect top-down processing, and may correspond to the subjective sense of unease characteristic of the uncanny valley effect. Objective neural components might be used in future to rapidly train generative artificial intelligence systems to produce more realistic images that are perceived as natural by human observers.

@@

Many people report an aversion to entities that are superficially human-like, but on closer inspection turn out to be artificial. Examples include humanoid robots (androids), puppets, hyper-realistic masks [@Sanders2021], and computer-generated images or movies. The term 'uncanny valley' [@Mori1970; English translation in @Mori2012] describes the idea that clearly human or clearly artificial entities do not evoke unease, whereas artificial entities that are human-like are disconcerting. Understanding these experiences is increasingly important as artificial entities become more integrated into our everyday lives. However at present relatively little is known about the neural underpinnings of the uncanny valley effect [for a recent review, see @Vaitonyte2023].

Neural responses to faces and bodies in general are well-characterised, and there appear to be specialised brain regions devoted to both [reviewed in @Hu2020]. For example, areas of the occipital lobe [@Gauthier2000] and fusiform gyrus [@Kanwisher1997] respond more to faces than non-face stimuli, and sections of extrastriate cortex are responsive to bodies [@Downing2001]. There are also event-related potential (ERP) signals associated with face and body stimuli, though their precise role is still debated [@Thierry2007; @Hong2022]. It seems highly likely that 'uncanny' images will activate these same processes, yet it is unclear whether the sense of unease they produce occurs at bottom-up sensory stages or is modulated by more top-down cognitive factors.

One previous study by @Saygin2012 measured fMRI responses to moving stimuli designed to elicit an uncanny valley effect. They found repetition suppression effects^[Repetition suppression is a phenomenon in which the neural responses to repeated presentations of identical or similar stimuli is reduced relative to the response on the first presentation.] in action-specific brain regions responding to movies of androids that had a biological appearance, but mechanical motion. These effects were stronger than for movies of humans or mechanical robots performing the same actions. A more recent electroencephalography (EEG) study [@Urgen2018] identified a difference in the N400 component^[The N400 is an electromagnetic brain potential obtained 400ms after stimulus onset, typically over centro-parietal electrodes. It has been proposed to reflect the extent to which the stimulus presented was surprising or unexpected.] between dynamic and static conditions using the same stimuli. Although this difference was strongest over frontal electrodes, source reconstruction of the N400 itself suggested a left-lateralised source in the temporo-parietal cortex, consistent with the fMRI results [@Saygin2012]. The authors interpret both of these findings as being due to the discrepancy between the human-like appearance and the clearly non-biological motion of the robot.

Our aim was to further investigate neural correlates of the uncanny valley effect. We achieved this through two EEG experiments, in which we measured neural responses to static images. In the first experiment, the stimuli were still images of humans, machine-like robots, and human-like robots (see Figure \ref{fig:stimexamples}a). In the second experiment we aimed to generalise the finding by using images of people wearing no masks, wearing obvious masks (e.g. Halloween masks), and wearing hyper-realistic silicone masks [@Sanders2017] (see Figure \ref{fig:stimexamples}b). Rather than focus on specific ERP components, we use a multivariate pattern classification approach to identify time windows in which information in the EEG signal can be used to distinguish between pairs of conditions. Our rationale is that timepoints where signals evoked by human faces can be distinguished from those evoked by human-like robots, or hyper-realistic masks, are candidates for a neural signature of the uncanny valley effect.

```{r stimexamples, fig.cap="Illustrative stimuli from the same categories as used in Experiments 1 and 2. Row (a) shows a human face, an android and a robot, all against white backgrounds. Row (b) shows a human face, a hyper-realistic silicone mask, and a Halloween mask, all against natural backgrounds. Images were taken from a variety of sources that permit reuse in academic contexts and in most cases were not part of the stimulus set from the experiments. The silicone mask image was taken by the authors, and was used in Experiment 2.", fig.align="center", out.width='80%', echo=FALSE}

@@

A total of 29 participants completed Experiment 1 (12 male, 17 female), and 30 different participants completed Experiment 2 (7 male, 23 female). None of the participants had previously taken part in a study using these stimuli, and all were naïve to the hypotheses and wore their normal optical correction if required. Written informed consent was collected before each experiment began, and all procedures were approved by the Ethics committee of [**redacted for blind peer review**].

@@

In Experiment 1, the stimulus set consisted of a total of 90 images, evenly split between three categories: real faces, human-like robots, and mechanical robots. Images all showed the head and shoulders of the subject, had white backgrounds, and were sourced from the internet. In Experiment 2, the stimulus set (first described by @Sanders2019, but here including additional images) consisted of a total of 296 images, comprising real faces (148 images), people wearing silicone masks (74 images), and people wearing obvious masks of the sort typically worn for carnivals and Halloween celebrations (74 images). The backgrounds of these images were more heterogeneous, and showed the natural surroundings of the subject. In both experiments, images involved examples of both genders, and of varied ethnic backgrounds.

@@

### Experiment 1: Robots

@@

After the EEG experiment, participants also completed a series of questionnaires using the Qualtrics platform. These involved rating their perception of a subset of the stimuli (8 from each category), using items from the Godspeed questionnaire [@Bartneck2009]. Inspection time was unlimited. Participants also provided demographic information (age, gender) and completed the GAToRS [@Koverola2022] and AQ [@Baron-Cohen2001] questionnaires, however the results of these additional questionnaires are not presented here.

### Experiment 2: Hyper-realistic masks

Participants were shown all 296 images in a random order in each of three blocks. In the first block, stimuli subtended $5.5\times7.5$ degrees of visual angle when viewed at a distance of 57cm. In the second block, stimuli doubled in size (width and height), and subtended $11\times15$ degrees at the same viewing distance. In the third block, stimuli doubled in size again, and subtended $22\times30$ degrees. The rationale for the size manipulation was to investigate whether increasing levels of detail made the silicone masks more identifiable [@Sanders2017; @Sanders2019]. However as that is not the main focus of the current paper, and our preliminary analyses indicated no differences between size conditions, we collapse results across size conditions. Stimuli were presented for 250ms, and participants indicated whether they thought each image contained a real face or a mask, using a two-button trackball. The button assignment (whether the left button indicated a face or a mask, and vice versa) was determined randomly for each participant, but remained constant throughout the whole experiment. Text reminding the participant of the button assignment was present continuously in the lower right corner of the screen, far from the area of the screen where the stimuli were presented. A central fixation cross was also present throughout. After each response there was a random duration blank period with a mean duration of 1000ms and a standard deviation of 200ms. Each block lasted around 8 minutes.

An independent group of 20 participants also completed an online questionnaire in which they rated the images along various dimensions. The participants repeated the real face vs mask judgement from the main experiment, and were additionally asked to rate emotional expressiveness, realism, and uncanniness for each image using a 7-point Likert scale. Inspection time was unlimited for these judgments.

@@

EEG signals were recorded during each block and saved to disc for subsequent offline analysis. We used components of the EEGlab toolbox [@Delorme2004] to import the data into Matlab and collate data across blocks. We then used Brainstorm [@Tadel2011] to filter the data using a bandpass filter (0.5 to 30Hz), epoch by condition, and subtract a pre-trial baseline (the mean voltage for the 200ms before stimulus onset). Five participants were excluded from the EEG analysis of each experiment due to excessive noise. Our attempts to clean up the data from these participants using independent components analysis were unsuccessful. Their behavioural data were unaffected, and therefore still included in the analysis.

@@

Multivariate pattern analysis was conducted by training a linear support vector machine algorithm [LibSVM, @Chang2011] to discriminate between patterns of activity across electrodes at a specific time point. The patterns came from the human face condition and one of the other conditions, for a single participant. Four examples of each pattern were calculated by averaging over random subsets of 20\% of the available trials from a given condition, and these were used to train the classifier. The accuracy of the classifier was tested on the remaining trials (that were not used in training) for each condition. This process was repeated 1000 times with different trial permutations to obtain an average accuracy, where chance performance is at 50\% correct. The analysis was carried out at all time points, and for each participant separately. We then averaged classifier accuracy across participants, and calculated one sample Bayesian t-tests comparing to chance performance at each time point as described above.

@@

Raw data, processed data, and analysis scripts are freely available through the project repository at: [**redacted for blind peer review**] 
<!-- https://doi.org/10.17605/OSF.IO/5NZ2H -->

@@

```{r exp1questionnaires, include=FALSE, results='hide'}

qlist <- c('Fake-Natural','Unconscious-Conscious','Apathetic-Responsive','Dislike-Like','Unfriendly-Friendly','Stagnant-Lively','Unintelligent-Intelligent','Irresponsible-Responsible','Anxious-Relaxed','Surprised-Quiescent')

robotQdata <- read.csv('local/robotquestionnairelong.csv')

filelist <- unique(robotQdata$File)
plist <- unique(robotQdata$Participant)
measurelist <- unique(robotQdata$Measure)

sortedQdata <- array(0,dim=c(length(plist),length(filelist),length(measurelist)))
for (file in 1:length(filelist)){
  conddata <- subset(robotQdata,File==filelist[file],)
  for (measure in 1:length(measurelist)){
    measuredata <- subset(conddata,Measure==measurelist[measure])
    for (p in 1:length(plist)){
      pdata <- subset(measuredata,Participant==plist[p])

      sortedQdata[p,file,measure] <- pdata$Response
    }
  }
}

conditionQdata <- array(0,dim=c(29,3,10))
conditionQdata[,1,] <- apply(sortedQdata[,1:8,],c(1,3),mean)
conditionQdata[,2,] <- apply(sortedQdata[,9:16,],c(1,3),mean)
conditionQdata[,3,] <- apply(sortedQdata[,17:24,],c(1,3),mean)

meanQdata <- apply(conditionQdata,c(2,3),mean)

pdf(paste0('Figures/RobotRatings.pdf'), bg="transparent", height = 30, width = 12)

par(mfrow=c(5,2))

plotlims <- c(1,5,0.5,3.5)  
ticklocsy <- c(0.5,1,2,3,3.5)    
ticklocsx <- seq(1,5,1)    # locations of tick marks on y axis
ticklabelsy <- c("","Robot","Android","Human","")       
ticklabelsx <- ticklocsx    # set labels for y ticks

for (cond in 1:10){
plot(x=NULL,y=NULL,axes=FALSE, ann=FALSE, xlim=plotlims[1:2], ylim=plotlims[3:4])  
axis(1, at=ticklocsx, tck=0.01, lab=F, lwd=2)     
axis(2, at=ticklocsy, tck=0.01, lab=F, lwd=2)
mtext(text = ticklabelsx, side = 1, at=ticklocsx,line=1,cex=1.5)     
mtext(text = ticklabelsy, side = 2, at=ticklocsy, line=0.2, las=3,cex=1.5) 
title(xlab=qlist[cond], col.lab=rgb(0,0,0), line=2.4, cex.lab=2)

text(1.15,3.3,paste0('(',letters[cond],')'),cex=3)

a <- density(conditionQdata[,1,cond],cut=6)
a$y <- 0.35*a$y/max(a$y)
polygon(a$x,1.1+a$y,border=NA,col=addalpha(collist[1],0.5))

a <- density(conditionQdata[,2,cond])
a$y <- 0.35*a$y/max(a$y)
polygon(a$x,2.1+a$y,border=NA,col=addalpha(collist[2],0.5))

a <- density(conditionQdata[,3,cond])
a$y <- 0.35*a$y/max(a$y)
polygon(a$x,3.1+a$y,border=NA,col=addalpha(collist[3],0.5))

xvals <- seq(-0.1,0.1,length=nrow(conditionQdata))
points(conditionQdata[,1,cond],1+xvals-0.2,pch=16,col=collist[1])
points(conditionQdata[,2,cond],2+xvals-0.2,pch=15,col=collist[2])
points(conditionQdata[,3,cond],3+xvals-0.2,pch=18,col=collist[3])

for (cond2 in 1:3){
  bs <- NULL
  for (n in 1:nbootstraps){bs[n] <- mean(sample(conditionQdata[,cond2,cond],replace=TRUE))}
  ci <- quantile(bs,c(0.025,0.975))
  arrows(mean(conditionQdata[,cond2,cond]),cond2,ci[1],cond2,angle=90,length=0.1,lwd=3)
  arrows(mean(conditionQdata[,cond2,cond]),cond2,ci[2],cond2,angle=90,length=0.1,lwd=3)
}

points(mean(conditionQdata[,1,cond]),1,pch=21,cex=2,lwd=2,bg=collist[1])
points(mean(conditionQdata[,2,cond]),2,pch=22,cex=2,lwd=2,bg=collist[2])
points(mean(conditionQdata[,3,cond]),3,pch=23,cex=2,lwd=2,bg=collist[3])

@@

EEG activity showed a clear visually evoked potential over posterior electrode sites (see Figure \ref{fig:RobotsData}c), with typical components found in response to visual stimuli pooled across all conditions (the P100, N170 and P200 are indicated in the figure). Pairwise comparisons of conditions are shown in Figure \ref{fig:RobotsMVPA}a,b. In general there is a tendency for the ERP response to human faces to diverge slightly from the other two conditions [@Geiger2021], however the evidence for this divergence was not compelling. Bayes factors exceeded 3 for only a small number of time points around 300-400ms in the comparison between human and robot images (see yellow bars at y = -8 in Figure \ref{fig:RobotsMVPA}a), but these differences were small considering the variance in the data.

@@

Finally, we analysed the rating data from a set of 10 questionnaire items for 8 stimuli from each category. The results are summarised in Figure \@ref(fig:robotratings), and in general show differences between stimulus categories along most dimensions. Of particular note, the U-shaped function predicted by the uncanny valley effect was apparent for ratings along the Dislike-Like (Figure \@ref(fig:robotratings)d), Unfriendly-Friendly (Figure \@ref(fig:robotratings)e), and Anxious-Relaxed (Figure \@ref(fig:robotratings)i) dimensions. These are all dimensions with emotional valence, indicating support for the 'uncanniness' of our android stimuli. However we note that the android and robot categories were typically rated as being more similar to each other than to the real human faces.

```{r robotratings, fig.cap="Ratings of stimuli using items from the Godspeed questionnaire. Each rating was on a Likert scale from 1-5, and was the average of ratings from 8 stimulus examples. Dots show individual participant scores, with the larger symbols indicating the mean and 95\\% bootstrapped confidence intervals.", fig.align="center", out.width="50%", echo=FALSE}

knitr::include_graphics('Figures/RobotRatings.pdf')

@@

```{r exp2questionnaires, include=FALSE, results='hide'}

maskQdata <- read.csv('local/maskquestionnairelong.csv')
comparisons <- matrix(c(1,2,1,3,2,3),nrow=3,ncol=2,byrow=TRUE)

hits <- matrix(0,nrow=20,ncol=3)
misses <- matrix(0,nrow=20,ncol=3)

accdata <- subset(maskQdata,maskQdata$Measure=='Acc')
for (s in 1:20){
  thissubj <- subset(accdata,accdata$Participant==s)
  condition <- NULL
  for (i in 1:nrow(thissubj)){
    if (grepl('hm',thissubj$File[i])){condition[i] <- 1}
    if (grepl('sm',thissubj$File[i])){condition[i] <- 2}    
    if (grepl('rf',thissubj$File[i])){condition[i] <- 3}
  }
  for (i in 1:nrow(thissubj)){
    hits[s,condition[i]] <- hits[s,condition[i]] + thissubj$Response[i] 
    misses[s,condition[i]] <- misses[s,condition[i]] + (1-thissubj$Response[i])
  }}

hitrate <- hits / (hits + misses)
falsealarms <- 0*hitrate
falsealarms[,1] <- misses[,3] / (misses[,3] + hits[,3])
falsealarms[,2] <- misses[,3] / (misses[,3] + hits[,3])
falsealarms[,3] <- rowMeans(misses[,1:2]) / (rowMeans(misses[,1:2]) + rowMeans(hits[,1:2]))

subjdprime <- qnorm(hitrate) - qnorm(falsealarms)
subjdprime[which(is.infinite(subjdprime))] <- 5
meandprimes <- colMeans(subjdprime)

condlist <- unique(maskQdata$Condition)
plist <- unique(maskQdata$Participant)
measurelist <- unique(maskQdata$Measure)

averageQdata <- array(0,dim=c(length(plist),length(condlist),length(measurelist)))

for (cond in 1:length(condlist)){
  conddata <- subset(maskQdata,Condition==condlist[cond],)
  for (measure in 1:length(measurelist)){
    measuredata <- subset(conddata,Measure==measurelist[measure])
    for (p in 1:length(plist)){
      pdata <- subset(measuredata,Participant==plist[p])

      averageQdata[p,cond,measure] <- mean(pdata$Response)
    }
  }
}
meanQscores <- apply(averageQdata,2:3,mean)

dataforANOVA <- melt(subjdprime)
colnames(dataforANOVA) <- c('Participant','Condition','dprime')
dataforANOVA$Participant <- as.factor(dataforANOVA$Participant)
dataforANOVA$Condition <- as.factor(dataforANOVA$Condition)
bfdprime2 <- extractBF(anovaBF(dprime ~ Condition,dataforANOVA,whichRandom='Participant'))$bf

bfcontrasts2 <- NULL
for (comp in 1:3){
bfcontrasts2[comp] <- extractBF(ttestBF(subjdprime[,comparisons[comp,1]],subjdprime[,comparisons[comp,2]],paired=TRUE))$bf}

dataforANOVA <- melt(averageQdata[,,1])
colnames(dataforANOVA) <- c('Participant','Condition','dprime')
dataforANOVA$Participant <- as.factor(dataforANOVA$Participant)
dataforANOVA$Condition <- as.factor(dataforANOVA$Condition)
bfEmoE <- extractBF(anovaBF(dprime ~ Condition,dataforANOVA,whichRandom='Participant'))$bf

bfcontrastsEmoE <- NULL
for (comp in 1:3){
bfcontrastsEmoE[comp] <- extractBF(ttestBF(averageQdata[,comparisons[comp,1],1],averageQdata[,comparisons[comp,2],1],paired=TRUE))$bf}

dataforANOVA <- melt(averageQdata[,,2])
colnames(dataforANOVA) <- c('Participant','Condition','dprime')
dataforANOVA$Participant <- as.factor(dataforANOVA$Participant)
dataforANOVA$Condition <- as.factor(dataforANOVA$Condition)
bfReal <- extractBF(anovaBF(dprime ~ Condition,dataforANOVA,whichRandom='Participant'))$bf

bfcontrastsReal <- NULL
for (comp in 1:3){
bfcontrastsReal[comp] <- extractBF(ttestBF(averageQdata[,comparisons[comp,1],2],averageQdata[,comparisons[comp,2],2],paired=TRUE))$bf}

dataforANOVA <- melt(averageQdata[,,3])
colnames(dataforANOVA) <- c('Participant','Condition','dprime')
dataforANOVA$Participant <- as.factor(dataforANOVA$Participant)
dataforANOVA$Condition <- as.factor(dataforANOVA$Condition)
bfUncan <- extractBF(anovaBF(dprime ~ Condition,dataforANOVA,whichRandom='Participant'))$bf

bfcontrastsUncan <- NULL
for (comp in 1:3){
bfcontrastsUncan[comp] <- extractBF(ttestBF(averageQdata[,comparisons[comp,1],3],averageQdata[,comparisons[comp,2],3],paired=TRUE))$bf}

pdf(paste0('Figures/MaskRatings.pdf'), bg="transparent", height = 12, width = 12)

plotlims <- c(1,7,0.5,3.5)  
ticklocsy <- c(0.5,1,2,3,3.5)    
ticklocsx <- seq(1,7,1)    # locations of tick marks on y axis
ticklabelsy <- c("","Halloween","Silicone","Human","")       
ticklabelsx <- ticklocsx    # set labels for y ticks

plot(x=NULL,y=NULL,axes=FALSE, ann=FALSE, xlim=plotlims[1:2], ylim=plotlims[3:4])  
axis(1, at=ticklocsx, tck=0.01, lab=F, lwd=2)     
axis(2, at=ticklocsy, tck=0.01, lab=F, lwd=2)
mtext(text = ticklabelsx, side = 1, at=ticklocsx,line=0.5,cex=1.5)     
mtext(text = ticklabelsy, side = 2, at=ticklocsy, line=0.2, las=3,cex=1.5) 
title(xlab="Emotional Expressiveness", col.lab=rgb(0,0,0), line=1.8, cex.lab=1.8)

text(1.15,3.3,'(a)',cex=3)

a <- density(averageQdata[,1,1],cut=6)
a$y <- 0.35*a$y/max(a$y)
polygon(a$x,1.1+a$y,border=NA,col=addalpha(collist[1],0.5))

a <- density(averageQdata[,2,1])
a$y <- 0.35*a$y/max(a$y)
polygon(a$x,2.1+a$y,border=NA,col=addalpha(collist[2],0.5))

a <- density(averageQdata[,3,1])
a$y <- 0.35*a$y/max(a$y)
polygon(a$x,3.1+a$y,border=NA,col=addalpha(collist[3],0.5))

xvals <- seq(-0.1,0.1,length=nrow(averageQdata))
points(averageQdata[,1,1],1+xvals-0.2,pch=16,col=collist[1])
points(averageQdata[,2,1],2+xvals-0.2,pch=15,col=collist[2])
points(averageQdata[,3,1],3+xvals-0.2,pch=18,col=collist[3])

for (cond in 1:3){
  bs <- NULL
  for (n in 1:nbootstraps){bs[n] <- mean(sample(averageQdata[,cond,1],replace=TRUE))}
  ci <- quantile(bs,c(0.025,0.975))
  arrows(mean(averageQdata[,cond,1]),cond,ci[1],cond,angle=90,length=0.1,lwd=3)
  arrows(mean(averageQdata[,cond,1]),cond,ci[2],cond,angle=90,length=0.1,lwd=3)
}

points(mean(averageQdata[,1,1]),1,pch=21,cex=2,lwd=2,bg=collist[1])
points(mean(averageQdata[,2,1]),2,pch=22,cex=2,lwd=2,bg=collist[2])
points(mean(averageQdata[,3,1]),3,pch=23,cex=2,lwd=2,bg=collist[3])

plotlims <- c(1,7,0.5,3.5)  
ticklocsy <- c(0.5,1,2,3,3.5)    
ticklocsx <- seq(1,7,1)    # locations of tick marks on y axis
ticklabelsy <- c("","Halloween","Silicone","Human","")       
ticklabelsx <- ticklocsx    # set labels for y ticks
